---
title: Chow’s Test
author: Benjamin Moran
date: '2017-11-17'
slug: chow-s-test
categories:
  - Flutracking
tags: 
  - Flutracking
---

One of the goals of my project is to try and identify changes in the structure of the various flutracking time series (initially the % of unvaccinated respondents reporting ILI symptoms). There are two reasons for this that inspire two separate lines of inquiry: does the structure change from year to year; and does the structure change within each year?

Here we will be using Chow's test to test for structural breaks in the flutracking data. Specifically, we'll be testing the % of unvaccinated respondents reporting influenza-like illness (ILI) symptoms using the `strucchange` package in `R`. Firstly, let's plot the %ILI Unvaccinated time series.

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE)
opts_chunk$set(warnings = FALSE)
library(tidyverse)
library(lubridate)
library(reshape2)
library(htmlTable)
library(strucchange)
```

```{r data, include = FALSE}
df <- readRDS("~/R/HonoursBlog/data/fludata.RDS")
df <- df %>%
  mutate(Year = factor(year(as.Date(.$Week_end)))) %>%
  mutate(Unvaccinated = .$Responses-.$Vaccinated_Respondents)

df2 <- melt(df,id = c("Week_end","fileref","Year"))
breaks <- match(unique(df$Year), df$Year)
labbreaks <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  labbreaks[i] = ((breaks[i+1]-breaks[i])/2) + breaks[i]
  labbreaks[j] = ((nrow(df)-breaks[j])/2) + breaks[j]
}
labelsbreak <- levels(df[breaks, "Year"])
cols <- scales::brewer_pal(palette = "Set1")(8)
nsurv <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  nsurv[i] = breaks[i+1]-breaks[i]
  nsurv[j] = nrow(df)-breaks[j]
}
firstsurv <- df[breaks, "Week_end"]
lastsurv <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  lastsurv[i] = df[(breaks[i]+nsurv[i]-1), 1]
  lastsurv[j] = df[nrow(df), "Week_end"]
}
lastsurv <- levels(droplevels(df[lastsurv,1]))

sumdat <- data.frame(Year = unique(df$Year),
                     Number = nsurv,
                     FirstSurvey = firstsurv,
                     LastSurvey = lastsurv)

endpoints <- sumdat[,2] + breaks - 1
endpoints[length(endpoints)] <-  nrow(df)
endpoints <- c(breaks, endpoints)

```


## Unvaccinated Repsondents Data

```{r dataTotal, include=FALSE}
cols <- scales::brewer_pal(palette = "Set1")(8)
dfresp <- df2[df2$variable %in% c("ILI_Unvaccinated"),] %>%
  group_by(variable) 
dfrespvals <- dfresp[endpoints,]
dfrespvals <- cbind(dfrespvals, breaks = endpoints,col = rep(factor(c("Start","End")),each = nrow(dfrespvals)/2))
dfrespvals %>% select(-c(fileref, variable)) -> dfrespvals

dfvline <- as.data.frame(rep(levels(droplevels(unique(dfresp$variable))),length(breaks))) %>%
  cbind(rep(breaks,each = 4))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)
```

```{r plotTotal}
ggplot() + 
  geom_path(data = dfresp,
            aes(x = Week_end,
                y = value,
                group = 1),
            colour = "#1f78b4",
            show.legend = FALSE) +
  geom_vline(data = dfvline,
             aes(xintercept = breaks), 
             colour = "lightgrey", 
             linetype = 2) +
  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df2$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.justification=c(1,0), 
        legend.position=c(0.95,0.8),
        legend.background = element_rect(colour = "lightgrey")) +
      geom_point(data = dfrespvals,
                 aes(x = Week_end,
                     y = value,
                     group = col,
                     colour = col)) + 
  scale_colour_manual(name  ="Survey Period",
                      breaks=c("Start", "End"),
                      labels=c("Start", "End"),
                      values = c("#33a02c","#e31a1c")) +
  labs(
    title = "Total Responses to the Flutracking Survey",
    subtitle = "by Year and survey week",
    caption = "Data from flutracking.net"
  ) 

```

Now I'll use the strucchange package to test for breaks. To test for breaks we fit a basic linear model: `lm(ILIts ~ 1)` where `ILIts` is the time series plotted abouve.

```{r strucchangeByYear}
ILIts <- ts(dfresp$value)
lvfit <- lm(ILIts ~ 1)
summary(lvfit)

ggplot() + 
  geom_path(data = dfresp,
            aes(x = Week_end,
                y = value,
                group = 1),
            colour = "#1f78b4",
            show.legend = FALSE) +
  geom_vline(data = dfvline,
             aes(xintercept = breaks), 
             colour = "lightgrey", 
             linetype = 2) +
  geom_hline(yintercept = lvfit$coefficients[1]) +
  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df2$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.justification=c(1,0), 
        legend.position=c(0.95,0.8),
        legend.background = element_rect(colour = "lightgrey")) +
      geom_point(data = dfrespvals,
                 aes(x = Week_end,
                     y = value,
                     group = col,
                     colour = col)) + 
  scale_colour_manual(name  ="Survey Period",
                      breaks=c("Start", "End"),
                      labels=c("Start", "End"),
                      values = c("#33a02c","#e31a1c")) +
  labs(
    title = "Total Responses to the Flutracking Survey",
    subtitle = "by Year and survey week",
    caption = "Data from flutracking.net"
  ) 


bp <- breakpoints(ILIts~ 1, h= 0.1)
summary(bp)
plot(bp)
```

So the breakpoints function suggests that there are 4 (min. BIC). 


```{r scplot2}

# iliwin <- window(ILIts, end = 185)
# 
# plot(iliwin, 
#      main = "%ILI w/ fit and breaks",
#      xaxt = "n",
#      xlab = NA,
#      ylab = "%ILI",
#      ylim = c(1,6),
#      lwd = 2)
# axis(side = 1, 
#      at = labbreaks, 
#      labels = labelsbreak,
#      tck=-.02)
# abline(v = breaks, col = "lightgrey", lty = 1)
# lines(confint(bp, breaks = 4))
# lines(fitted(bp, breaks = 4), col = 4, lwd = 1.5)
# 
# legend("topright", 
#        inset=c(.005,.01), 
#        cex = 0.7,
#        ncol = 1,
#        title="Legend", 
#        c("%ILI","Fit","Breaks","Conf.Int"), 
#        horiz=FALSE, 
#        lty=c(1,1,2,1), 
#        lwd=c(2,1.5,1,1), 
#        col=c("black","blue","black","red"))

bpfit <- data.frame(x = seq(1,185),values = c(fitted(bp, breaks = 4)))
bpconfint <- data.frame(confint(bp, breaks = 4)$confint,breaks = seq(1,4))
dfresp <- cbind(dfresp, week = seq(1,185))

ggplot() + 
  geom_path(data = dfresp,
            aes(x = week,
                y = value,
                group = 1),
            colour = "#1f78b4",
            show.legend = FALSE) +
  geom_vline(data = dfvline,
             aes(xintercept = breaks), 
             colour = "lightgrey", 
             linetype = 2) +
  geom_line(data = bpfit, aes(x= x, y = values)) +
  geom_segment(data = bpconfint, aes(x = X2.5.., xend = X97.5.., y = 1, yend = 1)) +
  geom_segment(data = bpconfint, aes(x = X2.5.., xend = X2.5.., y = 0.95, yend = 1.05)) +
  geom_segment(data = bpconfint, aes(x = X97.5.., xend = X97.5.., y = 0.95, yend = 1.05)) +
  geom_vline(data = bpconfint, aes(xintercept = breakpoints), lty = 2) +
  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df2$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.justification=c(1,0), 
        legend.position=c(0.95,0.8),
        legend.background = element_rect(colour = "lightgrey")) +
      geom_point(data = dfrespvals,
                 aes(x = breaks,
                     y = value,
                     group = col,
                     colour = col)) + 
  scale_colour_manual(name  ="Survey Period",
                      breaks=c("Start", "End"),
                      labels=c("Start", "End"),
                      values = c("#33a02c","#e31a1c")) +
  labs(
    title = "%ILI Reported by Unvaccinated Respondents",
    subtitle = "by Year and survey week",
    caption = "Data from flutracking.net"
  ) 

```

Here is the same data faceted by Year.

```{r dataByYear, include=FALSE}
dfresp2 <- cbind(dfresp, week = rep(NA,nrow(dfresp)))
dfresp2$week[[1]] = 1

for(i in 2:nrow(dfresp2)){
  if(dfresp2$Year[[i-1]] == dfresp2$Year[[i]]){
    dfresp2$week[[i]] = dfresp2$week[[i-1]] + 1
  } else {
    dfresp2$week[[i]] = 1
  }
}

dfresp2$week <- factor(dfresp2$week)
```


```{r plotByYear}
ggplot() + 
  geom_path(data = dfresp2,
            aes(x = week,
                y = value,
                group = 1),
            colour = "#1f78b4",
            show.legend = FALSE) +
  theme_minimal() +
  facet_wrap(~Year, 
             ncol = 3,
             scales = "free_x") +
  # scale_x_discrete(labels = labelsbreak, 
                   # breaks = df2$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(colour = "darkgrey"),
        axis.ticks.x = element_line(colour = "grey"),
        legend.justification=c(1,0), 
        legend.position=c(0.95,0.8),
        legend.background = element_rect(colour = "lightgrey")) +
  labs(
    title = "% ILI reported by Unvaccinated Respondents",
    subtitle = "by Year and Survey Week",
    caption = "Data from flutracking.net"
  ) 
```
