---
title: Summary of available time series
author: Benjamin Moran
date: '2017-11-14'
categories:
  - Flutracking
tags:
  - Flutracking
slug: available-time-series
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(reshape2)
library(htmlTable)

df <- readRDS("~/R/HonoursBlog/data/fludata.RDS")

# df <- readRDS("../../data/fludata.RDS") # work location
df %>%
  mutate(Year = factor(year(as.Date(.$Week_end)))) %>%
  mutate(Unvaccinated = .$Responses-.$Vaccinated_Respondents) %>%
  mutate(Clinical_Staff_Unvaccinated = .$Clinical_Staff-.$Clinical_Staff_Vaccinated) -> df

df2 <- melt(df,id = c("Week_end","fileref","Year"))
```


The data ranges from 2010 - 2017 but it is not an unbroken time series. The survey runs for about six months each year - usually beginning in May and ending in October - however the earlier years tend to have fewer surveys and each year does not span the same date range. The table below is a brief summary of this information.

# Summary Data

```{r}
breaks <- match(unique(df$Year), df$Year)
labbreaks <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  labbreaks[i] = ((breaks[i+1]-breaks[i])/2) + breaks[i]
  labbreaks[j] = ((nrow(df)-breaks[j])/2) + breaks[j]
}
labelsbreak <- levels(df[breaks, "Year"])
cols <- scales::brewer_pal(palette = "Set1")(8)
nsurv <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  nsurv[i] = breaks[i+1]-breaks[i]
  nsurv[j] = nrow(df)-breaks[j]
}
firstsurv <- droplevels(df[breaks, "Week_end"])
lastsurv <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  lastsurv[i] = df[(breaks[i]+nsurv[i]-1), 1]
  lastsurv[j] = df[nrow(df), "Week_end"]
}
lastsurv <- factor(df[lastsurv,1])

sumdat <- data.frame(Year = unique(df$Year),
                     Number = nsurv,
                     FirstSurvey = firstsurv,
                     LastSurvey = lastsurv)

# sumdat %>%  kable("html") %>%  kable_styling(bootstrap_options = c("striped", "hover"))
htmlTable(sumdat)
```


# Plots of Respondent Data

Here you can find a number of explanatory plots of the different time series available in the full data set.

## Total Respondents 

Below is plot of the total number of respondents to the Flutracking survey for each week. 

```{r Total}
cols <- scales::brewer_pal(palette = "Set1")(8)
dfresp <- df2[df2$variable %in% c("Responses"),]
dfvline <- as.data.frame(rep(levels(droplevels(unique(dfresp$variable))),length(breaks))) %>%
  cbind(rep(breaks,each = 4))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)

ggplot(data = dfresp,
       aes(x = Week_end, y = value, colour = Year, group = variable)) + 
  geom_point() +
  geom_path() +
  scale_colour_brewer(palette = "Set1") + 
  geom_vline(data = dfvline, aes(xintercept = breaks, colour = "lightgrey"), linetype = 2) +
  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "none",
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  labs(
    title = "Total Responses to the Flutracking Survey",
    subtitle = "by Year and survey week",
    caption = "Data from flutracking.net"
  ) 

```

## Types of Responses

Survey participants can respond both for themselves (_Self_) and for other (_Other_) members of their household. The below plot breaks the total response data down into these groups.

```{r}
cols <- scales::brewer_pal(palette = "Set1")(8)
dfresp <- df2[df2$variable %in% c("Self_Report", "Other_Report"),]
endpoints <- sumdat[,2] + breaks - 1
endpoints[length(endpoints)] <-  nrow(df)
endpoints <- c(breaks, endpoints)
dfrespvals <- dfresp[c(endpoints,(endpoints + nrow(dfresp)/2)),]
dfrespvals <- cbind(dfrespvals, rep(breaks,2))

dfvline <- as.data.frame(rep(levels(droplevels(unique(dfresp$variable))),2)) %>%
  cbind(breaks = rep(breaks,each = 2))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)

ggplot(data = dfresp,
       aes(x = Week_end, y = value, group = interaction(Year, variable))) + 
    geom_vline(data = dfvline, aes(xintercept = breaks),
             colour = "lightgrey",
             linetype = 2,
             show.legend = FALSE) +
  geom_path(aes(colour = variable)) +
  geom_point(data = dfrespvals,
             aes(x = Week_end,
                 y = value,
                 group = interaction(Year, variable),
                 colour = variable),
             show.legend = FALSE) + 
  scale_colour_discrete(name="Report Type",
                      breaks=c("Self_Report", "Other_Report"),
                         labels=c("Self", "Other")) +

  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  labs(
    title = "Total Responses to the Flutracking Survey",
    subtitle = "by Report Type",
    caption = "Data from flutracking.net"
  ) 

```


## Vaccination Status

Here the data is broken down into whether or not the subject of the survey (i.e. both self reports and others) has received the annual influenza vaccination. Note the spikes at the beginning of each year in the number of unvaccinated respondents due to the survey period beginning before some people have the chance to receive the vaccination. 

```{r SelfReport}
cols <- scales::brewer_pal(palette = "Set1")(8)
subgroups <- c("Vaccinated_Respondents","Unvaccinated")
dfresp <- df2[df2$variable %in% subgroups,]
endpoints <- sumdat[,2] + breaks - 1
endpoints[length(endpoints)] <-  nrow(df)
endpoints <- c(breaks, endpoints)
endpoints[6] <- endpoints[6] + 1
dfrespvals <- dfresp[c(endpoints,(endpoints + nrow(dfresp)/2)),]
dfrespvals <- cbind(dfrespvals, rep(breaks,2))

dfvline <- as.data.frame(rep(unique(dfresp$variable),2)) %>%
  cbind(breaks = rep(breaks,each = length(subgroups)))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)

ggplot(data = dfresp,
       aes(x = Week_end, y = value, group = interaction(Year, variable))) + 
    geom_vline(data = dfvline, aes(xintercept = breaks),
             colour = "lightgrey",
             linetype = 2,
             show.legend = FALSE) +
  geom_path(aes(colour = variable)) +
  geom_point(data = dfrespvals,
             aes(x = Week_end,
                 y = value,
                 group = interaction(Year, variable),
                 colour = variable),
             show.legend = FALSE) + 
  scale_colour_discrete(name="Vaccination Status",
                      breaks=c("Vaccinated_Respondents", "Unvaccinated"),
                         labels=c("Vaccinated", "Unvaccinated")) +

  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  labs(
    title = "Total Responses to the Flutracking Survey",
    subtitle = "by Vaccination Status",
    caption = "Data from flutracking.net\nNote: Vaccination data unnavailable for the first survey week of 2015 due to system error"
  )  
```

## Clinical Staff

```{r ClinicalStaff}
cols <- scales::brewer_pal(palette = "Set1")(8)
subgroups <- c("Clinical_Staff","Clinical_Staff_Vaccinated","Clinical_Staff_Unvaccinated")
dfresp <- df2[df2$variable %in% subgroups,]
# endpoints <- sumdat[,2] + breaks - 1
# endpoints[length(endpoints)] <-  nrow(df)
# endpoints <- c(breaks, endpoints)
# endpoints[6] <- endpoints[6] + 1
dfrespvals <- dfresp[dfresp$Week_end %in% firstsurv | dfresp$Week_end %in% lastsurv,]

dfvline <- as.data.frame(rep(unique(dfresp$variable),2)) %>%
  cbind(breaks = rep(breaks,each = length(subgroups)))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)


ggplot(data = dfresp,
       aes(x = Week_end, y = value, group = interaction(Year, variable))) + 
    geom_vline(data = dfvline, aes(xintercept = breaks),
             colour = "lightgrey",
             linetype = 2,
             show.legend = FALSE) +
  geom_path(aes(colour = variable)) +
  geom_point(data = dfrespvals,
             aes(x = Week_end,
                 y = value,
                 group = interaction(Year, variable),
                 colour = variable),
             show.legend = FALSE) + 
  scale_colour_discrete(name="Vaccination Status",
                      breaks=c("Clinical_Staff","Clinical_Staff_Vaccinated","Clinical_Staff_Unvaccinated"),
                         labels=c("Total","Vaccinated", "Unvaccinated")) +

  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  labs(
    title = "Total Responses to the Flutracking Survey - Clinical Staff",
    subtitle = "by Vaccination Status",
    caption = "Data from flutracking.net\nNote: Vaccination data unnavailable for the first survey week of 2015 due to system error"
  )  
```


# Plots of Influenza-like-illness Symptoms Data

## % ILI Reported

```{r ILIReported}
cols <- scales::brewer_pal(palette = "Set1")(8)
subgroups <- c("ILI_Vaccinated","ILI_Unvaccinated")
dfresp <- df2[df2$variable %in% subgroups,]
# endpoints <- sumdat[,2] + breaks - 1
# endpoints[length(endpoints)] <-  nrow(df)
# endpoints <- c(breaks, endpoints)
# endpoints[6] <- endpoints[6] + 1
dfrespvals <- dfresp[dfresp$Week_end %in% firstsurv | dfresp$Week_end %in% lastsurv,]

dfvline <- as.data.frame(rep(unique(dfresp$variable),2)) %>%
  cbind(breaks = rep(breaks,each = length(subgroups)))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)


ggplot(data = dfresp,
       aes(x = Week_end, y = value, group = interaction(Year, variable))) + 
    geom_vline(data = dfvline, aes(xintercept = breaks),
             colour = "lightgrey",
             linetype = 2,
             show.legend = FALSE) +
  geom_path(aes(colour = variable)) +
  geom_point(data = dfrespvals,
             aes(x = Week_end,
                 y = value,
                 group = interaction(Year, variable),
                 colour = variable),
             show.legend = FALSE) + 
  scale_colour_discrete(name="Vaccination Status",
                      breaks=c("ILI_Vaccinated","ILI_Unvaccinated"),
                         labels=c("Vaccinated", "Unvaccinated")) +

  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  labs(
    title = "% Influenza-like-illness Symptoms reported",
    subtitle = "by Vaccination Status",
    caption = "Data from flutracking.net"
  )  
```

## % ILI Reported with Absence

```{r ILIReportedwAbsence}
cols <- scales::brewer_pal(palette = "Set1")(8)
subgroups <- c("ILI_wAbsence_Vaccinated","ILI_wAbsence_Unvaccinated")
dfresp <- df2[df2$variable %in% subgroups,]
# endpoints <- sumdat[,2] + breaks - 1
# endpoints[length(endpoints)] <-  nrow(df)
# endpoints <- c(breaks, endpoints)
# endpoints[6] <- endpoints[6] + 1
dfrespvals <- dfresp[dfresp$Week_end %in% firstsurv | dfresp$Week_end %in% lastsurv,]

dfvline <- as.data.frame(rep(unique(dfresp$variable),2)) %>%
  cbind(breaks = rep(breaks,each = length(subgroups)))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)


ggplot(data = dfresp,
       aes(x = Week_end, y = value, group = interaction(Year, variable))) + 
    geom_vline(data = dfvline, aes(xintercept = breaks),
             colour = "lightgrey",
             linetype = 2,
             show.legend = FALSE) +
  geom_path(aes(colour = variable)) +
  geom_point(data = dfrespvals,
             aes(x = Week_end,
                 y = value,
                 group = interaction(Year, variable),
                 colour = variable),
             show.legend = FALSE) + 
  scale_colour_discrete(name="Vaccination Status",
                      breaks=c("ILI_wAbsence_Vaccinated","ILI_wAbsence_Unvaccinated"),
                         labels=c("Vaccinated", "Unvaccinated")) +

  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom",
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  labs(
    title = "% Influenza-like-illness Symptoms reported with an Absence from Work",
    subtitle = "by Vaccination Status",
    caption = "Data from flutracking.net"
  )  
```
