---
title: Chow’s Test 
author: Benjamin Moran
date: '2017-11-17'
slug: chow-s-test
categories:
  - Flutracking
tags: 
  - Flutracking
---

# Testing for changes in level

One of the goals of my project is to try and identify changes in the structure of the various flutracking time series (initially the % of unvaccinated respondents reporting ILI symptoms). There are two reasons for this that inspire two separate lines of inquiry: does the structure change from year to year; and does the structure change within each year?

Here we will be using Chow's test to test for structural breaks in the flutracking data from year to year. Specifically, we'll be testing the % of unvaccinated respondents reporting influenza-like illness (ILI) symptoms using the `strucchange` package in `R`. Firstly, let's plot the %ILI Unvaccinated time series.

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE)
opts_chunk$set(warnings = FALSE)
library(tidyverse)
library(lubridate)
library(reshape2)
library(htmlTable)
library(strucchange)
```

```{r data, include = FALSE}
df <- readRDS("~/R/HonoursBlog/data/fludata.RDS")
#df <- readRDS("../../data/fludata.RDS") # work location
df <- df %>%
  mutate(Year = factor(year(as.Date(.$Week_end)))) %>%
  mutate(Unvaccinated = .$Responses-.$Vaccinated_Respondents)

df2 <- melt(df,id = c("Week_end","fileref","Year"))
breaks <- match(unique(df$Year), df$Year)
labbreaks <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  labbreaks[i] = ((breaks[i+1]-breaks[i])/2) + breaks[i]
  labbreaks[j] = ((nrow(df)-breaks[j])/2) + breaks[j]
}
labelsbreak <- levels(df[breaks, "Year"])
cols <- scales::brewer_pal(palette = "Set1")(8)
nsurv <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  nsurv[i] = breaks[i+1]-breaks[i]
  nsurv[j] = nrow(df)-breaks[j]
}
firstsurv <- df[breaks, "Week_end"]
lastsurv <- c()
for(i in 1:(length(breaks)-1)){
  j = length(breaks)
  lastsurv[i] = df[(breaks[i]+nsurv[i]-1), 1]
  lastsurv[j] = df[nrow(df), "Week_end"]
}
lastsurv <- levels(droplevels(df[lastsurv,1]))

sumdat <- data.frame(Year = unique(df$Year),
                     Number = nsurv,
                     FirstSurvey = firstsurv,
                     LastSurvey = lastsurv)

endpoints <- sumdat[,2] + breaks - 1
endpoints[length(endpoints)] <-  nrow(df)
endpoints <- data.frame(breaks = c(breaks, endpoints))

```

## Unvaccinated Repsondents Data

```{r dataTotal, include=FALSE}
cols <- scales::brewer_pal(palette = "Set1")(8)

dfresp <- df2[df2$variable %in% c("ILI_Unvaccinated"),] %>%
  group_by(variable) 
endpoints <- sumdat[,2] + breaks - 1
endpoints[length(endpoints)] <-  nrow(df)
endpoints <- c(breaks, endpoints)
dfrespvals <- dfresp[c(endpoints),]
dfrespvals <- dfrespvals %>%
   mutate(col = rep(factor(c("Start","End")),each = nrow(dfrespvals)/2)) %>%
  mutate(breaks = endpoints)
# dfrespvals %>% select(-c(fileref, variable)) -> dfrespvals

dfvline <- as.data.frame(rep(levels(droplevels(unique(dfresp$variable))),length(breaks))) %>%
  cbind(rep(breaks,each = 1))

colnames(dfvline) <- c("variable","breaks")
dfvline <- group_by(dfvline, variable)
```

```{r plotTotal, ref.label="plotTotal"}
ggplot() + 
  geom_path(data = dfresp,
            aes(x = Week_end,
                y = value,
                group = 1),
            colour = "#1f78b4",
            show.legend = FALSE) +
  geom_vline(data = dfvline,
             aes(xintercept = breaks), 
             colour = "lightgrey", 
             linetype = 2) +
  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df2$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.justification=c(1,0), 
        legend.position=c(0.95,0.8),
        legend.background = element_rect(colour = "lightgrey")) +
      geom_point(data = dfrespvals,
                 aes(x = Week_end,
                     y = value,
                     group = col,
                     colour = col)) + 
  scale_colour_manual(name  ="Survey Period",
                      breaks=c("Start", "End"),
                      labels=c("Start", "End"),
                      values = c("#33a02c","#e31a1c")) +
  labs(
    title = "Total Responses to the Flutracking Survey",
    subtitle = "by Year and survey week",
    caption = "Data from flutracking.net"
  ) 

```

Now I'll use the strucchange package to test for breaks. To test for breaks we fit a basic linear model: `lm(ILIts ~ 1)` where `ILIts` is the time series plotted above.

```{r strucchangeByYear}
ILIts <- ts(dfresp$value)
lvfit <- lm(ILIts ~ 1)
summary(lvfit)

ggplot() + 
  geom_path(data = dfresp,
            aes(x = Week_end,
                y = value,
                group = 1),
            colour = "#1f78b4",
            show.legend = FALSE) +
  geom_vline(data = dfvline,
             aes(xintercept = breaks), 
             colour = "lightgrey", 
             linetype = 2) +
  geom_hline(yintercept = lvfit$coefficients[1]) +
  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df2$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.justification=c(1,0), 
        legend.position=c(0.95,0.8),
        legend.background = element_rect(colour = "lightgrey")) +
      geom_point(data = dfrespvals,
                 aes(x = Week_end,
                     y = value,
                     group = col,
                     colour = col)) + 
  scale_colour_manual(name  ="Survey Period",
                      breaks=c("Start", "End"),
                      labels=c("Start", "End"),
                      values = c("#33a02c","#e31a1c")) +
  labs(
    title = "Total Responses to the Flutracking Survey",
    subtitle = "by Year and survey week",
    caption = "Data from flutracking.net"
  ) 
```

Now we want to test to find significant departures from this basic model (intercept = `r summary(lvfit)$coefficients[[1]]`). We do this by calling the  `breakpoints` function with the basic `lm` as input.

```{r breakpoints, echo = c(1)}
bp <- breakpoints(ILIts~ 1, h= 0.1)
summary(bp)
```

So the breakpoints function suggests that there are 4 (min. BIC). We can confirm this by plotting the results of the model fitting.

```{r breakpoints2}
plot(bp)
```

The `breakpoints` function also gives us new fitted values for the intercept as well as confidence intervals for the breakpoints. We can use these to update \@ref(fig:plotTotal).

```{r scplot2}
bpfit <- data.frame(x = seq(1,185),values = c(fitted(bp, breaks = 4)))
bpconfint <- data.frame(confint(bp, breaks = 4)$confint,breaks = seq(1,4))
dfresp <- dfresp %>%
  mutate(week = seq(1,185))

ggplot() + 
  geom_path(data = dfresp,
            aes(x = week,
                y = value,
                group = 1),
            colour = "#1f78b4",
            show.legend = FALSE) +
  geom_vline(data = dfvline,
             aes(xintercept = breaks), 
             colour = "lightgrey", 
             linetype = 2) +
  geom_line(data = bpfit, aes(x= x, y = values)) +
  geom_segment(data = bpconfint, aes(x = X2.5.., xend = X97.5.., y = 1, yend = 1)) +
  geom_segment(data = bpconfint, aes(x = X2.5.., xend = X2.5.., y = 0.95, yend = 1.05)) +
  geom_segment(data = bpconfint, aes(x = X97.5.., xend = X97.5.., y = 0.95, yend = 1.05)) +
  geom_vline(data = bpconfint, aes(xintercept = breakpoints), lty = 2) +
  theme_minimal() +
  scale_x_discrete(labels = labelsbreak, 
                   breaks = df2$Week_end[labbreaks]) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.justification=c(1,0), 
        legend.position=c(0.95,0.8),
        legend.background = element_rect(colour = "lightgrey")) +
      geom_point(data = dfrespvals,
                 aes(x = breaks,
                     y = value,
                     group = col,
                     colour = col)) + 
  scale_colour_manual(name  ="Survey Period",
                      breaks=c("Start", "End"),
                      labels=c("Start", "End"),
                      values = c("#33a02c","#e31a1c")) +
  labs(
    title = "%ILI Reported by Unvaccinated Respondents",
    subtitle = "by Year and survey week",
    caption = "Data from flutracking.net"
  ) 

```

<!-- Here is the same data faceted by Year. -->

<!-- ```{r dataByYear, include=FALSE} -->
<!-- dfresp2 <- dfresp %>% -->
<!--   mutate(week = rep(NA,nrow(dfresp))) -->
<!-- dfresp2$week[[1]] = 1 -->

<!-- for(i in 2:nrow(dfresp2)){ -->
<!--   if(dfresp2$Year[[i-1]] == dfresp2$Year[[i]]){ -->
<!--     dfresp2$week[[i]] = dfresp2$week[[i-1]] + 1 -->
<!--   } else { -->
<!--     dfresp2$week[[i]] = 1 -->
<!--   } -->
<!-- } -->

<!-- dfresp2$week <- factor(dfresp2$week) -->
<!-- ``` -->


<!-- ```{r plotByYear} -->
<!-- ggplot() +  -->
<!--   geom_path(data = dfresp2, -->
<!--             aes(x = week, -->
<!--                 y = value, -->
<!--                 group = 1), -->
<!--             colour = "#1f78b4", -->
<!--             show.legend = FALSE) + -->
<!--   theme_minimal() + -->
<!--   facet_wrap(~Year,  -->
<!--              ncol = 3, -->
<!--              scales = "free_x") + -->
<!--   # scale_x_discrete(labels = labelsbreak,  -->
<!--                    # breaks = df2$Week_end[labbreaks]) + -->
<!--   theme(panel.grid.major.x = element_blank(), -->
<!--         panel.grid.minor.y = element_blank(), -->
<!--         axis.title.y = element_blank(), -->
<!--         axis.title.x = element_blank(), -->
<!--         axis.text.x = element_text(colour = "darkgrey"), -->
<!--         axis.ticks.x = element_line(colour = "grey"), -->
<!--         legend.justification=c(1,0),  -->
<!--         legend.position=c(0.95,0.8), -->
<!--         legend.background = element_rect(colour = "lightgrey")) + -->
<!--   labs( -->
<!--     title = "% ILI reported by Unvaccinated Respondents", -->
<!--     subtitle = "by Year and Survey Week", -->
<!--     caption = "Data from flutracking.net" -->
<!--   )  -->
<!-- ``` -->

<!-- Now I'll use the `breakpoints` function on data for each year individually. -->

<!-- ```{r} -->
<!-- # rm(bp) -->
<!-- # years <- unique(dfresp2$Year) -->
<!-- # for(i in 1:length(years)){ -->
<!-- #   x = years[i] -->
<!-- #   data = subset(dfresp2, Year == x) -->
<!-- #   assign(levels(years)[i], data) -->
<!-- # } -->

<!-- years <- unique(dfresp2$Year) -->
<!-- bpdata <- data.frame(Year = NA, bp1 = NA, bp2 = NA) -->
<!-- par(mfrow = c(3,3)) -->
<!-- for(i in 1:length(years)){ -->
<!-- x = subset(dfresp2, Year == years[i]) -->
<!-- data = breakpoints(x$value~1, h = 0.2) -->
<!-- bpdata[i,1] = levels(years)[i] -->
<!-- bpdata[i,2] = data$breakpoints[1] -->
<!-- bpdata[i,3] = data$breakpoints[2] -->
<!-- sumdata = summary(data)[["RSS"]][2,] -->
<!-- breaknum = as.integer(which(sumdata==min(sumdata))) -->
<!-- minBIC = round(min(summary(data)[["RSS"]][2,]),2) -->
<!-- plot(data, main = paste("Year:", years[i]),sub = paste0("min(BIC): ",minBIC), xlab = paste("Number of breakpoints", breaknum)) -->
<!-- } -->


<!-- ``` -->

<!-- ```{r plotByYear2} -->
<!-- ggplot() +  -->
<!--   geom_path(data = dfresp2, -->
<!--             aes(x = week, -->
<!--                 y = value, -->
<!--                 group = 1), -->
<!--             colour = "#1f78b4", -->
<!--             show.legend = FALSE) + -->
<!--   theme_minimal() + -->
<!--   facet_wrap(~Year,  -->
<!--              ncol = 3, -->
<!--              scales = "free_x") + -->
<!--   geom_vline(data = bpdata, aes(xintercept = bp1),  lty = 2, colour = "grey") + -->
<!--   geom_vline(data = bpdata, aes(xintercept = bp2),  lty = 2, colour = "grey") + -->
<!--   # scale_x_discrete(labels = labelsbreak,  -->
<!--                    # breaks = df2$Week_end[labbreaks]) + -->
<!--   theme(panel.grid.major.x = element_blank(), -->
<!--         panel.grid.minor.y = element_blank(), -->
<!--         axis.title.y = element_blank(), -->
<!--         axis.title.x = element_blank(), -->
<!--         axis.text.x = element_text(colour = "darkgrey"), -->
<!--         axis.ticks.x = element_line(colour = "grey"), -->
<!--         legend.justification=c(1,0),  -->
<!--         legend.position=c(0.95,0.8), -->
<!--         legend.background = element_rect(colour = "lightgrey")) + -->
<!--   labs( -->
<!--     title = "% ILI reported by Unvaccinated Respondents", -->
<!--     subtitle = "by Year and Survey Week", -->
<!--     caption = "Data from flutracking.net" -->
<!--   )  -->
<!-- ``` -->
