---
title: January Questions
author: Benjamin Moran
date: '2018-01-27'
slug: january-questions
categories:
  - Questions
tags:
  - Working Notes
---

```{r, include=FALSE}
library(knitr)
opts_chunk$set(eval = FALSE)
library(tidyverse)
library(lubridate)
library(reshape2)
library(htmlTable)
library(forecast)
df <- readRDS("~/R/HonoursBlog/data/fludata.RDS")

#df <- readRDS("../../data/fludata.RDS") # work location
df %>%
  mutate(Year = factor(year(as.Date(.$Week_end)))) %>%
  mutate(Unvaccinated = .$Responses-.$Vaccinated_Respondents) %>%
  mutate(Clinical_Staff_Unvaccinated = .$Clinical_Staff-.$Clinical_Staff_Vaccinated) -> df

df2 <- melt(df,id = c("Week_end","fileref","Year"))
```

I've reached a bit of an impasse and as a result I have a number of questions that I need to address, either with you or by myself.


## State-Space Models for Time Series Data

The methods that De Jong and Penzer use - and that the R packages I have used so far to reproduce some of their results - are reliant on representing the time-series as a state-space model in order to run the Kalman filter. Whilst De Jong and Penzer's paper does show variations of their method for ARIMA models (i.e. by representing them as a state-space model) they don't include a similar explanation for linear models like the one we arrived at last year.

So, the task I've given myself is to try and join the dots. I've got myself a textbook ('Time Series Analysis by State Space Methods' by J. Durbing and S.J. Koopman - it was sighted by virtually every package and R, as well as turning up in lots of other places) that I will go through to see if I can sort out my problem.

## Predictions

I've been trying to use some of R's prediction functions (`base::predict()` and `forecast::forecast()`) to see how the model performs, but I'm unsure how to proceed. Each of these methods requires that the components of the model that are used to predict are supplied in a $(m \times n)$ matrix, where $m$ is the number periods that we wish to predict and $n$ is the number of predictors.

Now, our model uses two different time series to predict what we're interested in: firstly, we use lags of the time series we are predicting; second, we use lags of another series. Whilst I could construct a loop that predicts each period individually before updating the data so that we always have access to the lags, we have no way of predict the other series.

Thus, unless we decide on some way to predict the other series (maybe `forecast::auto.arima()`), we couldn't use it as a predictor. 

```{r}
library(dynlm)
Unvacc <- ts(df$ILI_Unvaccinated,start = 1, end = nrow(df))
DF <- cbind(Unvacc = Unvacc,
            lag1 =  stats::lag(Unvacc, -1),
            lag2 =  stats::lag(Unvacc, -2))

DF <- data.frame(DF[1:185,])

fit <- tslm(Unvacc ~ stats::lag(Unvacc, 1)) # Model with 2 lags

n = 44
i = 1
for(i in 1:n){
  r = nrow(DF)
  newdata = data.frame(stats::lag(Unvacc, 1))
  x = forecast(fit,newdata)
  DF = rbind(DF, c(x,stats::lag(DF$Unvacc[r],-1),stats::lag(DF$Unvacc[r-1], -1)))
  i = i + 1
}

plot(ts(DF$Unvacc))
```


### De Jong and Penzer

### My Linear Model

There are two issues here. Firstly, I've worked myself into a state of confusion with regards to the model we arived at 

```{r}
Unvacc <- ts(df$ILI_Unvaccinated,start = 1, end = nrow(df))
Vacc <- ts(df$ILI_Vaccinated,start = 1, end = nrow(df))
DF <- cbind(Unvacc = Unvacc,
            lag1 =  stats::lag(Unvacc, -1),
            lag2 =  stats::lag(Unvacc, -2),
            Vacclag1 =  stats::lag(Vacc, -1))

DF <- data.frame(DF[1:185,])

# fit <- lm(Unvacc ~ lag1 + 0, data = DF) # Model with 1 lag
fit2 <- lm(Unvacc ~ lag1 + lag2 + 0, data = DF) # Model with 2 lags
# fit3 <- lm(Unvacc ~ lag1 + lag2 + Vacclag1 + 0, data = DF) # Model with 2 lags and Vacc data
# summary(fit)
summary(fit2)
# summary(fit3)
```


```{r}
Unvacc2 = data.frame(ID = seq(length(Unvacc)),Value = c(Unvacc))
fit1_res <- data.frame(ID = as.integer(names(fit$fitted.values))-1, Value = c(fit$fitted.values), res = c(fit$residuals))
fit2_res <- data.frame(ID = as.integer(names(fit2$fitted.values))-1, Value = c(fit2$fitted.values), res = c(fit2$residuals))
fit3_res <- data.frame(ID = as.integer(names(fit3$fitted.values))-1, Value = c(fit3$fitted.values), res = c(fit3$residuals))
ggplot(data = Unvacc2, aes(x = ID, y = Value)) + 
  geom_line() +
  # geom_line(data = fit1_res, aes(group = 2),col = "green") +
  geom_line(data = fit2_res,aes(group = 3),col = "red") + 
  geom_line(data = fit3_res,aes(group = 4),col = "blue")  

ggplot() + 
  geom_line(data = fit1_res, aes(group = 1,x = ID, y = res),col = "green") + 
  geom_line(data = fit2_res, aes(group = 2,x = ID, y = res),col = "red") + 
  geom_line(data = fit3_res, aes(group = 3,x = ID, y = res),col = "blue")
```

```{r}
Unvacc <- ts(df$ILI_Unvaccinated,start = 1, end = nrow(df))
Vacc <- ts(df$ILI_Vaccinated,start = 1, end = nrow(df))
DF <- cbind(Unvacc = Unvacc,
            lag1 =  stats::lag(Unvacc, -1),
            lag2 =  stats::lag(Unvacc, -2),
            Vacclag1 =  stats::lag(Vacc, -1))

DF <- subset(DF, start  = 0, end = 161)
fit <- lm(Unvacc ~ lag1, data = DF) # Model with 1 lag
h = 24
newdata <-
  cbind(Vacclag1 = rep(mean(as.numeric(DF[,4]), na.rm = TRUE), h),
        lag2 = rep(mean(as.numeric(DF[,3]), na.rm = TRUE), h),
        lag1=rep(mean(as.numeric(DF[,2]), na.rm = TRUE),h)) %>%
  rbind(c(0,0,0),.)%>%  as.data.frame()
fcast = forecast::forecast(fit, newdata = newdata)
fcast
```


The issue here is that I'm uncertain how to treat the linear model.


## Breaks in the Time Series

Interpolating missing values.

